Darwin's Dream: Unifying ecological and evolutionary change

### 2024-12-24 foundations_of_ecoevo.Rmd
Edits right before changing referencing style for nature

---
title: "Foundations of ecological and evolutionary change"
author: |
  | A. Bradley Duthie$^{1,a,*}$ and Victor J. Luque$^{2,a}$
date: '[1] Department of Biological and Environmental Sciences, University of Stirling, Stirling, Scotland [2] Department of Philosophy, University of Valencia, Valencia, Spain [*] Corresponding author: alexander.duthie@stir.ac.uk [a] Equal contribution'
output:
  pdf_document:
      keep_tex: yes
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{fancyhdr}
bibliography: refs.bib
link-citations: yes
linkcolor: blue
biblio-style: apalike
---

@@

@@

Rather than making simplifying assumptions, as is the approach for most ecological and evolutionary models, we focus on fundamental axioms that are universal to closed biological systems: discrete individuals, birth, death, and change over time.
We define an abstract sum ($\Omega$), to which all individuals within the population contribute.
From the simple assumptions of population identity ($z_{i} = 1$) and invariability ($\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 1) and principle of exponential growth ($N_{t+1} = \lambda N_{t}$).
By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution (Box 2).
We therefore deliver a general framework to unify fundamental equations of ecological and evolutionary change.

@@

<!---

Our unification and derivation of fundamental equations of ecology and evolution recovers and formalises well-known properties of biological systems.
The expressions $\beta_{i} - \delta_{i} + 1$ and $z_{i} - \Delta z_{i}$ can be interpreted to represent the ecological and evolutionary components of a system, respectively.
When focused only on change in population size, we set $z_{i} = 1$ and $\Delta z_{i} = 0$, then we can derive the fundamentals of population ecology without considering evolution.
Interestingly, the same is not true when focused only on evolutionary change.
If we set $\beta_{i} = 0$ and $\delta_{i} = 0$ (no birth or death), then there is no variance in fitness ($w_{i} = 1$ for all individuals), and evolution does not occur (note that evolution can still occur in a population at equilibrium where the birth rate and death rate are the same).
Fundamentally, evolutionary change is therefore enacted through ecological change, i.e., individual births and deaths, but ecological change does not necessarily imply evolution.
And while the link to population change is intuitive from the structure of eqn 1, the link to the Price equation is much less intuitive, especially after eqn 5 in our derivation.
The Price equation describes change in any evolving system, not just biological systems [@Fox2006; @Brantingham2007; @Bell2009a; @MacCallum2012; @Frank2015].
It is perhaps unsurprising that a derivation of ecological and evolutionary change from a common starting point requires relatively more labour to recover the Price equation.

--->

@@

<!---

PROBLEM HERE: TRYING TO RECOVER THE LOGISTIC GROWTH IN DISCRETE TIME

=======================================

Our fundamental equation is complete and exact.
It therefore implicitly includes any effects of density dependence in population growth (see Box 1).
Starting from eqn 1, we can partition ecological change into density-independent and density-dependent effects by multiplying the expression $\left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)$ within the summation by $\left(1 + \sum_{j=1}^{N} \alpha_{ij}z_{j}\right)$, where $j$ indexes the same individuals as $i$ and $\alpha_{ij}$ defines the effect $j$ has on the contribution of $i$ to $\Omega$. 
This partition can be used to recover

OR............. BELIEVE THAT THE BELOW IS THE CORRECT ONE!!

Our fundamental equation is complete and exact.
It therefore implicitly includes any effects of density dependence in population growth (see Box 1).
Starting from eqn 1, we can partition ecological change into density-independent and density-dependent effects by first defining $a_{ij}$ as the effect that the characteristic of $j$ ($z_{j}$) has on the contribution of $i$ to $\Omega$. 
Within the summation of eqn 1, we can then include a new term summing across all individuals in the population, $\sum_{j=1}^{N} a_{ij}z_{j}\right)$.
This sums the effect of each individual on a focal individual $i$.
If we define $\alpha = E\left( a_{ij}\right)$, then we can

--->

@@

<!---

NOTE 1
==================

Need to discuss that evolution *is* possible without ecology, if we imagine a set of RNA molecules that are all identical in an environment with no birth or death. We might have some change in RNA sequence (and therefore structure) that is evolution; this evolution will be in the \Delta z as non-zero -- the covariance of the Price equation will still be zero, but the expectation term will be nonzero because expected fitness is 1 (because you get credit for surviving from time step t to t+1). Hence, no adaptive evolution (selection) is possible without ecology, but mutation can happen (or selection at lower levels of biological organisation).

So evolution is possible without ecology and vice versa, however unlikely.

NOTE 2
================

We can partition \Delta z in the original equation however we want, and it just gets back to new terms attached to the Price equation at the end

Need to try to derive what happens when we partition fitness into birth, death, and survival component

--->

@@

>>>>

---
title: "Darwin's Dream: Unifying ecological and evolutionary change"
author: |
  | A. Bradley Duthie$^{1,a,*}$ and Victor J. Luque$^{2,a}$
date: '[1] Department of Biological and Environmental Sciences, University of Stirling, Stirling, Scotland [2] Department of Philosophy, University of Valencia, Valencia, Spain [*] Corresponding author: alexander.duthie@stir.ac.uk [a] Equal contribution'
output:
  pdf_document:
      keep_tex: yes
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 2
bibliography: refs.bib
link-citations: yes
linkcolor: blue
biblio-style: apalike
---

@@

**Key words:** Ecology, Evolution, Eco-Evolutionary Theory, Fundamental Theorem, Price Equation, Population Growth

Non-technical Summary
=====================================

Biologists have long known that the ecology of population size is inextricably linked with the evolution of population traits. Nevertheless, the foundational theory underlying ecological and evolutionary change have been historically separated, with recent attempts to bridge this theoretical divide using eco-evolutionary modelling. We discover a new equation from first principles that provides a universal and complete definition for eco-evolutionary change. This equation thereby offers a unified framework for ecology and evolution, which also recovers the foundational theory of ecology and evolution as developed in isolation.

@@

Rather than making simplifying assumptions, as is the approach for most ecological and evolutionary models, we focus on fundamental axioms that are universal to closed biological systems: discrete individuals, birth, death, and change over time.
We define an abstract sum ($\Omega$), to which all individuals within the population contribute.
From the simple assumptions of population identity ($z_{i} = 1$) and invariability ($\Delta z_{i} = 0$), we recover the most general equation of population ecology (Box 1) and principle of exponential growth ($N_{t+1} = N_{t}\lambda$).
By defining individual fitness ($w_{i}$) and applying the total conservation of probability to individual frequencies [@Frank2015; @Frank2016], we recover the most fundamental equation of evolution (Box 2).
We therefore deliver a general framework to unify fundamental equations of ecological and evolutionary change.

@@

@@

@@

@@

Author Contributions
====================================================

Both authors came up with the question idea. 
Following many discussions between the authors, ABD proposed the initial equation with subsequent exploration and development from both authors. 
Both authors contributed to the writing.

Competing Interests
====================================================

The authors declare no competing interests.

### 2024-12-24 foundations_of_ecoevo.tex
Edits right before changing referencing style for nature

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Foundations of ecological and evolutionary change},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fancyhdr}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Foundations of ecological and evolutionary change}
\author{A. Bradley Duthie\(^{1,a,*}\) and Victor J. Luque\(^{2,a}\)}
\date{{[}1{]} Department of Biological and Environmental Sciences,
University of Stirling, Stirling, Scotland {[}2{]} Department of
Philosophy, University of Valencia, Valencia, Spain {[}*{]}
Corresponding author:
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
{[}a{]} Equal contribution}

@@

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

@@

@@

Rather than making simplifying assumptions, as is the approach for most
ecological and evolutionary models, we focus on fundamental axioms that
are universal to closed biological systems: discrete individuals, birth,
death, and change over time. We define an abstract sum (\(\Omega\)), to
which all individuals within the population contribute. From the simple
assumptions of population identity (\(z_{i} = 1\)) and invariability
(\(\Delta z_{i} = 0\)), we recover the most general equation of
population ecology (Box 1) and principle of exponential growth
(\(N_{t+1} = \lambda N_{t}\)). By defining individual fitness
(\(w_{i}\)) and applying the total conservation of probability to
individual frequencies (\protect\hyperlink{ref-Frank2015}{Frank 2015},
\protect\hyperlink{ref-Frank2016}{2016}), we recover the most
fundamental equation of evolution (Box 2). We therefore deliver a
general framework to unify fundamental equations of ecological and
evolutionary change.

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Darwin's Dream: Unifying ecological and evolutionary change},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fancyhdr}
\usepackage{lineno}
\linenumbers
\modulolinenumbers[2]
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Darwin's Dream: Unifying ecological and evolutionary change}
\author{A. Bradley Duthie\(^{1,a,*}\) and Victor J. Luque\(^{2,a}\)}
\date{{[}1{]} Department of Biological and Environmental Sciences,
University of Stirling, Stirling, Scotland {[}2{]} Department of
Philosophy, University of Valencia, Valencia, Spain {[}*{]}
Corresponding author:
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
{[}a{]} Equal contribution}

@@

\setstretch{2}
\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

@@

\textbf{Key words:} Ecology, Evolution, Eco-Evolutionary Theory,
Fundamental Theorem, Price Equation, Population Growth

\hypertarget{non-technical-summary}{%
\section{Non-technical Summary}\label{non-technical-summary}}

Biologists have long known that the ecology of population size is
inextricably linked with the evolution of population traits.
Nevertheless, the foundational theory underlying ecological and
evolutionary change have been historically separated, with recent
attempts to bridge this theoretical divide using eco-evolutionary
modelling. We discover a new equation from first principles that
provides a universal and complete definition for eco-evolutionary
change. This equation thereby offers a unified framework for ecology and
evolution, which also recovers the foundational theory of ecology and
evolution as developed in isolation.

@@

Rather than making simplifying assumptions, as is the approach for most
ecological and evolutionary models, we focus on fundamental axioms that
are universal to closed biological systems: discrete individuals, birth,
death, and change over time. We define an abstract sum (\(\Omega\)), to
which all individuals within the population contribute. From the simple
assumptions of population identity (\(z_{i} = 1\)) and invariability
(\(\Delta z_{i} = 0\)), we recover the most general equation of
population ecology (Box 1) and principle of exponential growth
(\(N_{t+1} = N_{t}\lambda\)). By defining individual fitness (\(w_{i}\))
and applying the total conservation of probability to individual
frequencies (\protect\hyperlink{ref-Frank2015}{Frank 2015},
\protect\hyperlink{ref-Frank2016}{2016}), we recover the most
fundamental equation of evolution (Box 2). We therefore deliver a
general framework to unify fundamental equations of ecological and
evolutionary change.

@@

\hypertarget{author-contributions}{%
\section{Author Contributions}\label{author-contributions}}

Both authors came up with the question idea. Following many discussions
between the authors, ABD proposed the initial equation with subsequent
exploration and development from both authors. Both authors contributed
to the writing.

\hypertarget{competing-interests}{%
\section{Competing Interests}\label{competing-interests}}

The authors declare no competing interests.
